Search techniques for gaussian mixture modelling with Minimum Message Length, with applications to the origin of our Galaxy

### 2018-10-17 article/ms.tex
words

\documentclass{elsarticle}
\journal{Artificial Intelligence Journal}
\usepackage{bm,url,amsmath,amssymb}
%\usepackage{boldsymbol}
\usepackage{color,colortbl} % for \todo command
\usepackage{verbatim} % for \begin{comment}'ing out sections.
\usepackage{algorithm,algorithmicx}
\usepackage{algpseudocode}

@@

          \noindent{}and by making use of logarithmic identities:
          %  $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad . \label{eq:slw_nonuniform}
          \end{equation}
          %\begin{equation}
          %      \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
          %\end{equation}
\end{enumerate}

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
derived the bounds
%\begin{eqnarray}
%\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
%\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
%\end{eqnarray}
\begin{equation}
-K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
\end{equation}
\noindent{}which  describe the smallest and largest contribution that the weights can contribute to the message length for any untrialled mixture with $K$ components. When making predictions for future mixtures, we record the optimal weights for each trialled mixture and calculate a \emph{uniformity fraction} $f_u \in [0, 1]$
that describes the uniformity of the mixture from Equation \ref{eq:slw_nonuniform} ($f_u = 0$) to Equation \ref{eq:slw_uniform} ($f_u = 1$). We use estimates of $f_u$ from previously trialled mixtures to make predictions for future, untrialled mixtures.

@@

The \emph{Message-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of model evaluations.

\todo{initialisation, how kmeans++ breaks when too big unless covariance regularisation\\}

\todo{Run E-M, make predictions\\}

\todo{Advantage in being able to jump many $\Delta{}K$ at once}

@@

To evaluate the proposed search algorithms, we conduct a set of experiments on synthetic and real datasets.

@@

\todo{describe APOGEE data}

@@

\paragraph{Simulations}

\paragraph{Criteria for Comparison}

\todo{Convergence criteria: how often do we reach the `true' solution?}

\todo{What is the variance in the results of our method?}

\todo{Speed comparisons}

@@

\section{Appendices}
\subsection{The MML objective function}
\label{app:mml_derivation}
In the following subsections we provide expressions for the message length of
each component listed above, and our prior beliefs, before describing our
full objective function.

>>>>

\documentclass{elsarticle}
\journal{Artificial Intelligence Journal}
\usepackage{bm,url,amsmath,amssymb}
%\usepackage{boldsymbol}
\usepackage{color,colortbl} % for \todo command
\usepackage{verbatim} % for \begin{comment}'ing out sections.
\usepackage{algorithm,algorithmicx}
\usepackage{algpseudocode}
\usepackage{mathtools}

\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}

@@

          \noindent{}and by making use of logarithmic identities:
          %  $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
          %\begin{equation}
          %      \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + %\log{N} + \log{\left(N - 2K + 2\right)} \quad . \label{eq:slw_nonuniform}
          %\end{equation}
          \begin{equation}
            \sum_{k=1}^{K}\log\weight_k = (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
          \end{equation}

          %\begin{equation}
          %      \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
          %\end{equation}
\end{enumerate}

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
derived the bounds
%\begin{eqnarray}
%\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
%\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
%\end{eqnarray}
\begin{equation}
-K\log{K} \quad \geq \quad \sum_{k=1}^{K}\log\weight_k \quad \geq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
\end{equation}
\noindent{}which  describe the smallest and largest contribution that the weights can contribute to the message length for any untrialled mixture with $K$ components. When making predictions for future mixtures, we record the optimal weights for each trialled mixture and calculate a \emph{uniformity fraction} $f_u \in [0, 1]$
\begin{equation}
f_u = \frac{\alpha - \sum\log\weight_k}{\alpha + K\log{K}}
\end{equation}
\noindent{}where for brevity $\alpha = (K-1)\log{2} + \log(N-2K+2)-K\log{N}$,
and $f_u$ the uniformity of the mixture from Equation \ref{eq:slw_nonuniform} ($f_u = 0$) to Equation \ref{eq:slw_uniform} ($f_u = 1$). We use estimates of $f_u$ from previously trialled mixtures to make predictions for the uniformity of future,
untrialled mixtures, and the message length required to encode the weights.
Given the theoretical bounds in Equations \ref{eq:slw_uniform} and \ref{ew:slw_nonuniform},
we clip the predictions of the expectation value on the message length required to encode the weights by these bounds.

@@

The \emph{Message-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of computations. In the \emph{Message-Breaking Method} we
try to accomplish this using partial E-M to minimise the total number of computations
required to converge on a $K+1$ component mixture from a $K$ component mixture.
However, in the \emph{Message-Jumping Method} we seek to minimise the total number
of computations by optimizing a fewer number of mixtures. Instead of iteratively
perturbing a mixture from say $K = 10$ to $K = 30$, we use predictions of the
message length of untrialled mixtures and immediately `jump' to a component (
e.g., $K = 27$) that will provide the most improvement in our predictions of the
minimum message length over \emph{all} components. In doing so we seek a global 
message length minima with a fewer number of total E-M operations.

We describe the \emph{Message-Jumping Method} in Algorithm~\ref{alg:mbm}. In this
method we seek to predict the minimum message length for any number of $K$ components,
and then converge to the optimal mixture as quickly as possible. For this reason,
we initialise many mixtures that are evenly separated in $\log_{10}$ space from
$K = 1$ to $K = N/2$ in order to
quickly describe the minimum message length across all $K$. Specifically we
introduce a search hyperparameter $K_{inits}$ (defaults to 10) that sets the
number of mixtures to initialise -- and those initial $K$ values. 
We initialise each mixture using the K-means++ algorithm\cite{kmeanspp} and
optimize the mixture using EM as described earlier. We trial each of the $K_{init}$
mixtures in rank order of increasing $K$. If one of these mixtures fails to initialise
or optimise, this is likely because we could not perform Choleksy decomposition
of one (or more) of the covariance matrices. In practice this is typically because
the estimate of $K$ is much higher than $K_{true}$, leading to near-zero values
along the diagonal of the covariance matrix. In this situation we do not attempt
to initialise any more of the $K_{init}$ mixtures -- although higher $K$ values
can be trialled later -- and we being the search method.

The motivation for this initialisation process is to mimic the search strategy
that a human might take: What is an order-of-magnitude estimate on the number of
$K$? Or what is a ballpark number on the maximum number of $K$ that we should
trial? After we have that number, and an estimate on the message length for a
few sparsely-sampled numbers of components, we can make reasonable predictions
on the global minimum.

We record the message length for the optimal mixture at each trialled $K$, and
the model parameters for that mixture. With these quantities we predict the
message length of untrialled mixtures for every $K$ value from $K = 1$ to $K = \ceil{1.1\max(K_{trialled})}$ -- e.g., up to 10\% ahead of the largest previously trialled
$K$ value. The message length predicted from the optimised Gaussian processes
provide an expectation value on the message length for an untrialled mixture, as
well as a variance in that prediction (Figure \ref{fig:search-predictions}).
Before selecting the next value of $K$ to trial, we use the expectation value and
associated variance to compute the acquisition function (or the "expected
improvement") at each value of $K$ \cite{Jones:1998}.

We select the next $K$ value by choosing the value that will maximise the expected
improvement \todo{but this forces us out to large K, so ....}

\todo{initialisation with kmeans++, if it fails,....}

@@

We conducted a set of experiments and real datasets to evaluate the proposed
search algorithms.

@@

We use detailed chemical abundances from the Apache Point Observatory Galactic Evolution Experiment (APOGEE), an experiment run by the Sloan Digital Sky Survey (SDSS) collaboration. We make use of SDSS Data Release 12, which includes high-resolution, high S/N infrared spectra for $\sim$163,000 stars collected by APOGEE. Each observation
is an array of 8,575 pixels that have an associated rest-frame wavelength, and a
measurement of the stellar flux that is normalised relative to the spectral energy
distribution such that a flux value of 1 indicates no stellar absorption and a
flux value near 0 implies strong stellar absorption. All else being equal, a stronger
amount of absorption (lower normalised flux value) at a particular wavelength 
indicates that the star contains more of the chemical element that absorbs and
emits light at that wavelength. For example, sodium emits and absorbs light 
at X\,\AA\ when an electron moves between the XY transitions. All else being
equal, a star with lower normalised flux at a rest-frame wavelength of X\,\AA\
indicates that the star has a higher abundance of sodium.

We make use of the detailed chemical abundances calculated by \citep{casey}
from 96,648 APOGEE spectra. This data set includes 15 measured chemical abundances
for all stars (e.g., no missing data), with estimated uncertainties. All chemical
abundances are reported as the logarithmic count of the number of atoms relative
to hydrogen, with a correction to account for the Sun's chemical composition such
that "[Na/H] = 0" indicates a star has the same abundance of sodium relative to
hydrogen as the Sun. \todo{To provide context on the fidelity of the data, in Figure \ref{fig:gce} we show the measured abundance
of [Na/H] with respect to the measured abundance of [Fe/H] for all XX,XXX stars,
where the Sun's abundance is marked.}

The data set also contains other measured properties of the
stars, including their temperature and surface gravity, and whether they are 
likely associated with any gravitationally-bound star cluster. Most stars are
not likely members of any known star cluster; most stars in this dataset are 
described as `field' stars because they have no obvious association.
We do not make use of any of this ancillary data (e.g., temperature, cluster
association) during our search, but we do make use of it to evaluate the performance
of our search.

Costs of generated data sets.
Convergence of generated data sets.
When ther eis a switchover and we are faster.

Expectation that kasarapu and allison should be O(3) and we are O(1) with some overhead.

\begin{figure}
    \includegraphics[width=1.0\textwidth]{cost.pdf}
    \caption{Elapsed wall clock time for trialled search strategies to converge 
             on generated datasets as a function of the true number of clusters
             in the data. The approximate cost functions are shown.}
    \label{fig:cost}
\end{figure}

@@

@@

\section{Appendices}
\subsection{The MML objective function}
\label{app:mml_derivation}
In the following subsections we provide expressions for the message length to
encode a gaussian mixture model, and our prior beliefs on the parameters of that model, before describing our
full objective function.
